Handling String-Number Conversion in String Constraint Solving

### 2019-11-18 draft.tex
polish intro

Here $value_0$ and $last_0$ are string variables and others are integer variables. Line 1 describes the initial condition: \textsf{value} should be a sequence of digits and \textsf{sum} is initially zero. Lines 2-4 and lines 5-7 describe one execution of the first and second loop, respectively. Line 8 describes the condition of $i_2$ before leaving the loop. Finally, Lines 9-10 describe the condition that the last digit of sum is zero. Observe that to describe such a program path, we need a solver that supports the following types of constraints:
\begin{itemize}
	\item \emph{regular} constraints (e.g., $value_0 \in [0,9]^+$, which says $value_0$ is in the regular language $[0,9]^+$),
	\item \emph{integer} constraints (e.g., $i_0 = |value_0| -1$, which says $i_0$ equals the length of $value_0$ minus one),
	\item \emph{equality} constraints (often $y=\chatAt{x}{i}$ is encoded as $x=x_1.x_2.x_3 \wedge |x_1| = i \wedge |x_2| =1 \wedge y= x_2$, which uses equality of string terms $x$ and $x_1.x_2.x_3$), and
	\item \emph{string-number conversion} (e.g., $\its{sum_2}$, which is the string value of the number $sum_2$).
\end{itemize}

@@

\begin{verbatim}
function resetArray(arr) {
  for(var i = 0; i < arr.length; i++) {
    arr[i] = 0
  }
}
\end{verbatim}

A casual glance at the above code reveal no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: the arrays are indexed by strings, and numeric indexes are converted to strings. Moreover, the \textit{length} of an array is defined as the ``largest integer index plus one''. Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that uses {\tt{resetArray}}, with the value of {\tt{x}} shown on the right:

\begin{tabular}{l|c}
	{\tt{x = [0, 1, 2, 3, 4, 5]}} & [0, 1, 2, 3, 4, 5] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0] \\
	{\tt{x[3] = 5}} & [0, 0, 0, 5, 0, 0] \\
	{\tt{x["12"] = 7}} & [0, 0, 0, 5, 0, 0, , , , , , 7] \\
	{\tt{resetArray(x)}} & [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0] \\
\end{tabular}

In this code, string and integer values access the same array elements interchangeably, necessitating conversions between string and integers to faithfully model JavaScript semantics.  This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}).

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, falls into the case where neither the string nor the integer in a conversion is constant and hence most SMT solvers will not be able to handle it. More specifically, in some program the assignment {\tt{x[k] = 5}} can happen before calling {\tt{resetArray(x)}}, where {\tt{k}} is an user input string. One would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

@@

Solving string constraint with string-number conversion is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments, all of the state-of-the-art string constraint solvers do not support string-number conversion correctly. Some of them handles such constraints with only over-approximation of the original formula, and is known to have one-side error (see Section~\ref{section:evaluation} for detailed explanation).  \yfc{add concrete evidence}

>>>>

Here $value_0$ and $last_0$ are string variables and others are integer variables. Line 1 describes the initial condition: \textsf{value} should be a sequence of digits and \textsf{sum} is initially zero. Lines 2-4 and lines 5-7 describe one execution of the first and second loop, respectively. Line 8 describes the condition of $i_2$ before leaving the loop. Finally, Lines 9-10 describe the condition that the last digit of sum is zero. Observe that to describe such a program path, we need a solver that supports the following types of constraints:
\begin{itemize}
	\item \emph{regular} constraints (e.g., $value_0 \in [1,9]^+$, which says $value_0$ is in the regular language $[0,9]^+$),
	\item \emph{integer} constraints (e.g., $i_0 = |value_0| -1$, which says $i_0$ equals the length of $value_0$ minus one),
	\item \emph{equality} constraints (often $y=\chatAt{x}{i}$ is encoded as $x=x_1.x_2.x_3 \wedge |x_1| = i \wedge |x_2| =1 \wedge y= x_2$, which uses equality of string terms $x$ and $x_1.x_2.x_3$), and
	\item \emph{string-number conversion} (e.g., $\its{sum_2}$, which is the string value of the number $sum_2$).
\end{itemize}

@@

\begin{verbatim}
function resetArray(arr) {
  for(var i = 0; i < 10; i++) {
    arr[i] = 0
  }
}
\end{verbatim}

A casual glance at the above code reveal no use of strings at all, but the semantics of field access is somewhat unusual in JavaScript: the arrays are indexed by strings, and numeric indexes are converted to strings. This conversion is mandated explicitly by JavaScript semantics: the 2019 edition of ECMAScript requires that $ToPropertyKey$ be called on the element expression (\S{12.3.2.1}), and $ToPropertyKey$ calls {\tt{ToString}} on that value in all but special cases (\S{7.1.14}). Any faithful symbolic execution of JavaScript must handle such conversions for even basic array operations to work correctly; consider the following code snippet that manipulates an array \texttt{x}, with its value shown on the right:

\hspace{-4mm}\begin{tabular}{l|l|c}
1&	{\tt{x = [0, 0, 0, 0, 0]}} & [0, 0, 0, 0, 0] \\
2&	{\tt{x[3] = 4}} & [0, 0, 0, 4, 0] \\
3&	{\tt{x["3"] = 5}} & [0, 0, 0, 5, 0] \\
4&	{\tt{x["03"] = 7}} & [0, 0, 0, 5, 0] and x["03"] = 7\\
5&	{\tt{x["03"-1] = 2}} & [0, 0, 2, 5, 0] and x["03"] = 7\\
\end{tabular}

Here both \texttt{x[3]} in line 2 and \texttt{x["3"]} in line 3 denote the array element of \texttt{x["3"]}, but \texttt{x["03"]} denotes a completely different element. So naive idea of modeling array index using integers will not work -- it cannot distinguish the indexes \texttt{"3"} and \texttt{"03"}.

If we model array index with strings, then we need to face the second problem -- we need to handle arithmetics in the array index. Let us look at the case of line 4, we need to update the value of \texttt{x["03"-1]}. The expression \texttt{"03"-1} will be evaluated to $2$, which involves an implicit type conversion from the string \texttt{"03"} to an integer value $3$ due to the \texttt{-} operation. Then the number $2$ will be converted back to string \texttt{"2"} and used as array index and hence \texttt{x["03"-1]} is the array element of \texttt{x["2"]}. Even for a simple example like this, the conversion between string and number is unavoidable. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

\hide{

The Line 4 of the code updates the length of \texttt{x} to $13$ and hence the \texttt{resetArray(x)} will update \texttt{x} to an array with $13$ zeros.
Now suppose that the Line 4 of the code above is replace to {\tt{x[k] = 5}}, where $\mathtt{k}$ is an user input string and we want to find an input such that $\mathtt{x[15]=0}$. In this case, one would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. More precisely, in the corresponding string constraint encoding, we use a shadow variable $xLen$ to remember the length of $x$ and update its value after all array assignment. For an assignment $\mathtt{x[k] = v}$, we update the array value and length using the following string constraint (combined the $select$ function from SMT array theory): $x_{i+1} = store(x_i,k, v)$ and $xLen_{i+1} = ite( xLen_{i}> \sti{k}, xLen_{i}, \sti{k})$.
This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code.

\hide{In this code, string and integer values access the same array elements interchangeably, necessitating conversions between string and integers to faithfully model JavaScript semantics.

Now observe that {\tt{resetArray}} assigns to the array in a loop of which the size cannot, in general, be known a priori; as such, falls into the case where neither the string nor the integer in a conversion is constant and hence most SMT solvers will not be able to handle it. More specifically, in some program the assignment {\tt{x[k] = 5}} can happen before calling {\tt{resetArray(x)}}, where {\tt{k}} is an user input string. One would need to know the value of $\sti{k}$ and hence whether it will affect the size of the array {\tt{x}}. This is a rather straightforward array operation, and so this limit will cripple any model of non-trivial JavaScript code. }

@@

Solving string constraint with string-number conversion is a very challenging problem. From the theoretical point of view, this problem is already proven to be undecidable~\cite{day2018satisfiability}. In practice, in our experiments, the state-of-the-art string constraint solvers provide only limited support to string-number conversion.
